\documentclass[]{report}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage[margin=0.9in]{geometry}
\usepackage[toc,page]{appendix}
\usepackage{mathtools}
\renewcommand{\chaptername}{}
\lstdefinestyle{sharpc}{
 language=[Sharp]C, 		
 belowcaptionskip=1\baselineskip,
 breaklines=true,
 frame=L,
 xleftmargin=\parindent,
 showstringspaces=false,
 basicstyle=\footnotesize\ttfamily,
 rulecolor=\color{blue!80!black}
 }
    
% Title Page
\title{ECOTE Project (A21):  Program that converts C\# code with lambda expressions into code without lambda expressions}

\author{Author: Paweł Paczuski \\ Supervisor: dr inż. Anna Derezińska}

\begin{document}
    \maketitle
    
    \tableofcontents
    \pagebreak
    
    \chapter{Theoretical introduction}
    \section{Lambda calculus in Mathematics}
    Lambda calculus was introduced as a mean of formalizing the concept of expressing computation based on function abstraction and application using variable binding and substitution long before computers and programming languages existed. \cite{lambddef}
    The central concept in $\lambda$ calculus is "expression". A  "name",  also  called a "variable", is an identifier which, for our purposes, can be any of the letters
    a,b,c,...
    An expression is defined recursively as follows\cite{lambdintro}:\\
    $<$expression$>:=< $name$ >|< $function$ >|< $application$ > $ \\
    $<$function $>:= \lambda < $name$ >.< $expression$ > $\\
    $<$application$>:=<$expression$ >< $expression$ > $\\
    
    Lambda  calculus can be used to express any computable function. Because of this fact it is equivalent to Turing machines but the computational power of lambda calculus is not connected in any way with hardware implementation. This concept is more likely to be implemented using software.
    
    \subsection{Evaluations}
Function applications are evaluated by substituting the value of the argument x in the body of the function definition, i.e.
\begin{equation}
    (\lambda x.x)y=[y/x]x=y
\end{equation}
The notation $[y/x]$ is used to indicate that all occurrences of x are substituted by y in the expression to the right \cite{lambdintro}.
Another example of substitution:
\begin{equation}
(\lambda x.(\lambda y.(x(\lambda x.xy))))y
\end{equation}
In order not to mix variable y occurrences - it is used as a bound variable and free variable in this expression, it should be renamed\cite{lambdintro}.
\begin{equation}
[y/x](\lambda t.(x(\lambda x.xt))) = (\lambda t.(y(\lambda x.xy)))
\end{equation}
Lambda expression is evaluated until it is not possible to reduce it more.\cite{lambdintro}
    \section{Lambda calculus in programming}
    
    In spite of being object-oriented, languages C\#, JAVA and C++ borrow several concepts from functional paradigm. One of them is the concept of anonymous function. Exemplary situation in which anonymous functions are useful can be filtering array using some set of predicates. Existing array is being iterated over and a function is called with each element of the array as its argument. The function examines properties of the element and returns boolean value. If the function returns true, the element is being pushed to the array that is result of filtering the former one. Otherwise, this element is skipped. 
    Using purely object-oriented approach, one would have to create an implementation of an interface that has a function that returns boolean value and is able to decide whether an element of the array satisfies all conditions. The problem with this solution is the distance between code that makes decision about an element and the place where function that filters the array was called. When somebody reads code written in this manner, the have to go to the definition of class and take a look at implementation of the interface method. A lot of boilerplate code is introduced and namespace is polluted with simple classes that provide no special functionality.
    This is the place where traditional $\lambda$ calculus gives practical tool that increases language expressiveness and therefore allows for writing code that is easier to read and has more bugs. One could define an anonymous function that decides about filtering an item in the place of calling filter function making the code much more expressive. This idea was implemented in C\# language mechanism called LINQ 
    
    \section{Lambda expressions in C\#}\label{lambdatypes}
    Creators of C\# language decided to split syntax of a lambda expressions into several types.
    Lambda expressions in C\# language consist of arguments, $=>$ operator and expression or statement block.
    The $=>$ operator has the same precedence as assignment (=) and is right associative \cite{csharplambdasyntax}
    
    \subsection{Simple lambda expressions}
    The simplest lambda expressions. They are constituted of a single argument, $=>$ operator and expression. Types of arguments and returned value can be inferred. 
    \begin{lstlisting}[style=sharpc, caption=This lambda has single integer argument m and returns an integer m - 1]
    Func<int, int> lambd1 = m => m - 1;
    \end{lstlisting}

    \subsection{Parenthesized lambda expressions}    
    When lambda has more than one argument, parentheses have to be present. Single-argument lambda expression with parentheses also belongs to this category.
    \begin{lstlisting}[style=sharpc, caption=Lambda arguments can have types defined explicitly]
    Func<int, int, int> lambd2 = (int m, int n) => (m - n);
    \end{lstlisting}
    
    \subsection{Statement lambdas}
    More complex lambdas can contain statements within their body so, in addition to return value, they can have side effects.
    \begin{lstlisting}[style=sharpc, caption=execution lambd3 has side-effects that have no impact on the returned value]
    string text = "abc";
    Func<int, int> lambd3 = (n) =>
    {
        Console.WriteLine(text);
        Console.WriteLine(n);
        return n % 2;
    };
    \end{lstlisting}
    
    \subsection{Lambda expressions capturing local variables}
        Lambda expressions can implicitly capture local/instance variables. This has some serious implications e.g. when lambda expression capturing variable in a local block is assigned to a field of a class and then called in a different context, it has to be able to read value of a variable that used to be present in scope where it was assigned to the class field.
        \begin{lstlisting}[style=sharpc, caption="lambd4 has access to the local variable text in any context that. This property is sometimes called closure"]        
        string text = "this is some local text";
        Func<string, string> lambd4 = (n) => (text+n);
        \end{lstlisting}
    
    \subsection{Recursive lambda expressions}
    These are lambda expressions that can define other lambda expressions in their body. Due to limited practical applications (their syntax is not easy to read which is contrary to the expectation that lambdas are used to make code more readable), they are not described in a detailed way in this document.
    

        \begin{lstlisting}[style=sharpc, caption="factorial defined using lambda"]
        Func<Func<int, int>, Func<int, int>> factorial = (fac) => x => x == 0 ? 0 : x * fac(x - 1); 
        \end{lstlisting}
    
    \chapter{Lambda Converter for C\#}
    \section{Idea}
    The goal is to create a program (from now on referred to as Converter) that as an input receives source code written in C\# language that may contain lambda expressions of various kind. The program performs transformations of several kinds on the code in order to output modified version (but having the same functionality) of the code that contains no lambda expressions.
    \section{Transformation stages}
    The inner workings of the program can be split conceptually into several stages:
    \begin{enumerate}\label{transstages}
    \item       Syntactic analysis
    \item  		Semantic analysis
    \item       Code transformation
    \end{enumerate}
    \subsection{Syntactic analysis}
    In this stage the source code is analyzed in order to create syntax tree of the code. This data structure keeps all information about syntactic constructs in the input code. For a given syntax tree node, one can get information about its type, access modifier, identifier etc. 
    \subsection{Semantic analysis}
    Lambda conversion may require type inference for arguments and results, data flow information (what local scope variables are used in lambda's body), etc. These types of information can be taken from semantic model that is build using previously obtained syntax tree.
    
    \subsection{Code transformation}
    
    Information gathered in previous steps is used to find all lambda expressions in code that the program is capable of transforming. For each node transformation algorithm (described in section \ref{transalgorithm}) is applied.
    After all compatible lambda expressions have been replaced, the resulting syntax tree is being converted into code it represents, and resulting code is reformatted in order to increase readability.
    
    \section{Transformation algorithm}\label{transalgorithm}
    \begin{enumerate}
      \item {Determine which kind of lambda expression syntax described in section  \ref{lambdatypes} current lambda has}
   \item Determine names and types of lambda expression arguments
   \item Analyze what variables from the scope are referenced in lambda's body
   \item Create syntax node for a class declaration, give it unique identifier
   \item Add fields to the class syntax node that correspond to types of variables and names of variables referenced in lambda's body
   \item Create method syntax node that has the same return value as lambda, receives the same arguments and infer types of lambda's arguments. Its identifier can be e.g. "LambdaMethod". For statement lambda, copy original body into new method. For expression lambdas, add return keyword and concatenate with original body.
   \item Insert method syntax node into newly created class syntax node
   \item Insert class declaration syntax node into body of a class that contains lambda expression
   \item Find last expression or statement before original lambda expression.
   \item Create syntax nodes that correspond to instantiation of class, filling fields with values of variables from current scope. Insert those nodes just after the previously found node.
    \item Create method reference syntax node. Replace syntax node that used to define lambda expression with method reference node.
    \end{enumerate}
    
    \section{Roslyn}
    \subsection{Motivation}
    The problem of converting code containing lambda expressions has a lot of inherited complexity. The C\# language is a multi-paradigm language that evolved since its first release. Therefore, a grammar that defines it has many productions. The task of writing lexer and parser on our own as a subtask of the whole project would take a lot of time and effort to finish. Luckily, the creators of C\# language provide a set of open-source tools used to analyze and transform code called Roslyn. Roslyn exposes the C\#  compiler’s code analysis to a programmer as a consumer by providing an API layer that mirrors a traditional compiler pipeline. \cite{roslynrepo} Roslyn is used in the first two stages of operation of the Converter described in \ref{transstages}.

    \subsection{Architecture}
   
    Roslyn consists of two layers of API: compilers API and workspaces API.\\
    Compilers API provides programmer with object models that are used at each stage of the compilation: syntactic and semantic.\\
    Workspace API provides an object model to aggregate the code model across projects in a solution. Mainly for code analysis and refactoring around IDEs like Visual Studio, though the APIs are not dependent on Visual Studio\cite{workspaceapi}. 
    
    \subsubsection{Syntactic model}
    Syntactic model is represented mainly by a syntax tree whose nodes are called syntax nodes-constructs that represent syntactic elements such as statements, declarations, clauses, expressions, etc.\\ Implementation of syntax tree used in Roslyn is immutable, which means that any change to nodes of the tree results in creation of a clone of the tree with selected nodes different. This allows for parallelism as most of Roslyn APIs are thread-safe.
    
    \subsubsection{Semantic model}
    Semantic model adds to the syntactic model information about language rules and how the data flows in regions of source code. 
    Semantic model uses symbols as representations of a distinct element declared by the source code or imported from an assembly as metadata. Every name-space, type, method, property, field, event, parameter, or local variable is represented by a symbol\cite{roslynoverview}.
    
    \subsubsection{Workspace}
    Workspace is an abstraction that describes solution as a collection of projects that contain documents with source code. This API is used in Converter in order to make syntax node insertion easier and efficient.
    
    \section{Data structures used by Lambda Converter}   
    Main method of Program class contains code that is responsible for bootstrapping LambdaConverter and loading code from files that are passed to the program using command-line arguments. It uses File class to load contents of input file and save results to a file.
    
    The whole program's logic described in \ref{transalgorithm}  is implemented in  LambdaConverter class. This class provides static method Convert(string code) 
    
    The program uses basic container types provided by C\# language such as Lists and Dictionaries. \\Code analysis is performed using Roslyn's implementation of immutable syntax tree. Generation of C\# syntax elements is performed using SyntaxFactory\cite{syntaxFactory} abstraction. It allows for precise definition of syntax elements accepted by the language. 
    \\    For each syntax node that represents a lambda expression, a structure TransformationInfo is instantiated and is used to hold data: 
    \begin{itemize}
    \item Reference to original lambda syntax node;
    \item Syntax node that represents class declaration generated by the algorithm, to be inserted into the original code
    \item Syntax node that represents class instantiation 
    \item List of syntax nodes that represent and filling its fields just before passing LambdaMethod into the place where original lambda expression was defined
    \item Syntax node that represents passing instance of LambdaMethod to the place where lambda expression was defined
    \item Reference to the node after which code that instantiates the class and fills class should be put
    \end{itemize}
    
    
    \section{How to use Lambda Converter}
    The program requires no special configuration, it either can accept and convert the input code or it fails. The program is ran from console by specifying input file/s names. Files with converted code have names with .converted extension.
    
    \begin{lstlisting}[caption={Example of invocation of LambdaConverter. Input code is present in code.cs file. The output will be stored in code.cs.converted file}]
    lambda_converter.exe code.cs
    \end{lstlisting}

    \section{Limitations}
        The program was developed with assumption that the input is C\#code that would normally be stored in a single file. On the other hand, this assumption can be changed  quite easily to introduce support for more complex input.
        \\
     In spite of having the capability of capturing local variables to be used in lambda body, Converter lacks the functionality that would capture fields of a class in which lambda expression is defined. 
     \\
     Another limitation is no support for nested lambda expressions. This is a construction that is very  rarely used in practice, so it was decided that this language feature can be omitted in the proof-of-concept implementation.
    \\
    Some minor differences can be observed when one deals with C\# value types (float, double). They are internally implemented as structs named Single, Double respectively. This has no direct effects on the execution of output code.
    \section{Errors}
    Main class of errors that can occur while executing Converter are categorized into three groups:
    \begin{itemize}
        \item Improper input code. Program throws ImproperInputCodeException whenever a code that is no a proper C\# code is provided as input.
        \item Unsupported transformation. Program throws UnsupportedCodeTransformationException whenever input contains code has lambda expressions that can not be converted using the implemented algorithm. 
        \item EnvironmentError LambdaConverter can not get contents of the supplied file or write to output file.
    \end{itemize}

    
    \section{Tests}
This section contains description of scenarios that were taken into consideration while developing the program. Verifications of majority of them can be observed in the exemplary input/output sections\ref{egIO}.

     \begin{itemize}

         \item Test if SimpleLambdaExpressions are converted properly
         \item Test if ParenthesisedLambdaExpressions are converted properly
         \item Test if lambda expression containing statements are converted properly
         \item Test if lambda expressions inserted into LINQ constructs are converted properly
         \item Test if arguments of the lambda expressions are inferred
         \item Test if arguments with types specified are used properly
         \item Test if Converter captures all variables that are  used in lambda body
        \item Test if Converter signals an error when lambda captures field of the ancestor-class
        \item Test if Converter signals an error when input code is not a proper C\# code
        \item Test if Converter deals with conversion of reference types, value types, built-in types, user-defined  types
        \item Test if Converter instantiates generated classes properly, all fields are filled
        \item Test if Converter replaces lambda expression with method
        \item Test if Converter reformats the output code
        \item Test if all fields of generated class are public
        \item Test behavior of the program, when code that had been transformed, then had new lambda expressions added is once again transformed
        \item Test if all class names generated by the Converter are unique
        \item Test behavior of the program for input code taken from real-world projects-code is present in ExternalLambdaCode.cs file
        \end{itemize}
     

    
    
    
    \section{Performance considerations}
    Due to immutability of the syntax tree implementation provided by Roslyn, each insertion creates a new snapshot of the tree. This has to be taken into consideration when inserting/removing nodes. Solution to it is to use built-in mechanism of batch-insertions. Because each modification creates new tree, a remembered place in the code is lost. To keep track of changes DocumentEditor abstraction can be used. It is helpful when one deals with many modifications but requires usage of Workspace API that has performance drawbacks.
    
    \section{Practical applications}
    Increasing popularity of refactoring tools like JetBrain's Resharper or tools included in Visual Studio installation like Intellisense or refactor-suggestions make it quite possible that Lambda Converter could be used in this kind of tooling to allow for quick transformations of portions of code. Program in its current form could be quite easily converted into a Visual Studio extension.
    \\
    One could also use Lambda Converter to teach about lambda expressions and how they can be implemented in programming languages. This can be especially useful for teaching programmers that are proficient in object-oriented programming and want to expand their knowledge about functional  programming.


\section {Exemplary input and output}\label{egIO}
    \subsection{Input}\label{testsInput}
    \lstinputlisting[style=sharpc]{"bin/Debug/targetCode/LambdaCode.cs"}
    \pagebreak
    \subsection{Output}\label{testsOutput}
    \lstinputlisting[style=sharpc]{"bin/Debug/targetCode/LambdaCode.cs.converted"} 
    \pagebreak
 
                  
    
    
\begin{thebibliography}{80}
    \bibitem{lambddef} https://en.wikipedia.org/wiki/Lambda\_calculus accessed \date{01.05.2017}
    \bibitem{lambdintro} R. Rojas, A Tutorial Introduction to the Lambda Calculus
    \bibitem{csharpgrammar} https://msdn.microsoft.com/en-us/library/aa664812(v=vs.71).aspx
    \bibitem{csharplambdasyntax} https://msdn.microsoft.com/en-us/library/bb397687.aspx
    accessed \date{01.05.2017}
    \bibitem{roslynrepo} https://github.com/dotnet/roslyn
    accessed \date{01.05.2017}
    \bibitem{roslynoverview} https://github.com/dotnet/roslyn/wiki/Roslyn\_ Overview
    accessed \date{01.05.2017}
    \bibitem{workspaceapi} http://www.amazedsaint.com/2011/10/c-vnext-roslynan-introduction-and-quick.html
    accessed \date{02.05.2017}
\bibitem{syntaxFactory} http://source.roslyn.io/\#Microsoft.CodeAnalysis.CSharp/Syntax/SyntaxFactory.cs accessed \date{08.05.2017}

\end{thebibliography}
\end{document}